Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-10-12 paper.tex
small changes
updated figures+paper+bib

%NEW
\newcommand{\rmn}{\mathrm}
\newcommand{\bx}{{\bf x}}
\newcommand{\fg}{{F_\gamma}}
\newcommand{\sg}{{S_\gamma}}
\newcommand{\psf}{\theta_\rmn{res}}
\newcommand{\ph}{\rmn{ph}}
\newcommand{\eph}{E_\ph}
\newcommand{\nph}{N_\ph}
\newcommand{\vir}{\rmn{vir}}
\newcommand{\gal}{\rmn{gal}}
\newcommand{\sd}{\rmn{SD}}
\newcommand{\clu}{\rmn{cl}}
\newcommand{\sfe}{\rmn{sfe}}
\newcommand{\sub}{\rmn{sub}}
\newcommand{\msun}{M_\odot}
\newcommand{\ee}{E_\rmn{e}}
\newcommand{\stars}{\rmn{stars}}
\newcommand{\dust}{\rmn{dust}}
\newcommand{\lx}{L_\rmn{X}}
\newcommand{\s}{\rmn{s}}
\newcommand{\Kp}{\rmn{K}^\prime}
\newcommand{\Ip}{\rmn{I}^\prime}
\newcommand{\Js}{\rmn{J}^*}
\newcommand{\Jp}{\rmn{J}^\prime}
\newcommand{\B}{\rmn{B}}
\newcommand{\bsub}{\B_\rmn{sub}}
\newcommand{\qCR}{q_{\rmn{CR}-\ensuremath{\pi^0}}}
\newcommand{\ds}{{\sc DarkSUSY}}
\newcommand{\mlim}{M_\rmn{lim}}
\newcommand{\sm}{\rmn{sm}}
\newcommand{\mpc}{\rmn{Mpc}}
\newcommand{\kpc}{\rmn{kpc}}
\newcommand{\mev}{\rmn{MeV}}
\newcommand{\cm}{\rmn{cm}}
\newcommand{\egt}{\tilde{E}_\gamma}
\newcommand{\eet}{\tilde{E}_\rmn{e}}
\newcommand{\bra}{\langle}
\newcommand{\ket}{\rangle}
\newcommand{\vst}{\vspace{-0.14mm}}
\newcommand{\el}{\rmn{el}}

@@

\begin{abstract}
  We study the possibility for detecting gamma-ray emission from galaxy
  clusters. We consider 1) leptophilic models of dark matter (DM) annihilation
  that include a Sommerfeld enhancement (SFE), 2) different representative
  benchmark models of supersymmetric DM, and 3) cosmic ray (CR) induced pion
  decay. To model DM annihilation, we consider continuum emission as a result of
  the hadronization of annihilating neutralinos, internal bremsstrahlung, and
  inverse Compton emission from the cosmic microwave background as well as from
  a realistic spatial and spectral distribution of dust and stellar light. Among
  all clusters/groups of a flux-limited X-ray sample, we predict Virgo, Fornax
  and M49 to be the brightest DM sources and find a particularly low CR-induced
  background for Fornax. For a minimum substructure mass given by the DM
  free-streaming scale, cluster halos maximize the substructure boost for which
  we find a factor of $\gtrsim 1000$. Since regions around the virial radius
  dominate the annihilation flux of substructures, the resulting surface
  brightness profiles are almost flat. This makes it very challenging to detect
  this flux with imaging atmospheric Cherenkov telescopes since their
  sensitivity drops approximately linearly with radius and they typically have
  $5-10$ linear resolution elements across a cluster. Assuming cold dark matter
  with a substructure mass distribution down to an Earth mass and using extended
  \Fermi upper limits, we rule out the leptophilic models in their present form
  in 28 clusters, and limit the boost from SFE in M49 and Fornax to be $\lesssim
  5$. This corresponds to a limit on SFE in the Milky Way of $\lesssim 3$ which
  is too small to account for the increasing positron fraction with energy as
  seen by PAMELA. Alternatively, if SFE is realized in Nature, this would imply
  a limiting substructure mass of $\mlim > 10^4\,\msun$---a problem for
  structure formation in most particle physics models. Using individual cluster
  observations, it will be challenging for \Fermi to constrain our selection of
  DM benchmark models without SFE. The \Fermi upper limits are, however, closing
  in on our predictions for the CR flux using an analytic model based on
  cosmological hydrodynamical cluster simulations. We limit the CR-to-thermal
  pressure in nearby bright galaxy clusters of the \Fermi sample to $\lesssim
  10\%$ and in Norma and Coma to $\lesssim 3\%$. Thus we will soon start to
  constrain the underlying CR physics such as shock acceleration efficiencies or
  CR transport properties.
\end{abstract}

@@

Dark matter (DM) has been searched for in direct detection experiments
\cite{Pato:2010zk}, at accelerators
\cite{Ellis:2001hv,Baer:2006ff,Khachatryan:2011tk} and also in
indirect detection experiments looking for signals in the cosmic-ray
(CR) spectra of antiprotons, positrons, neutrinos and all of the
electromagnetic spectrum from radio waves to gamma-rays
\cite{Bergstrom:2009ib}. So far, the improvements in direct detection
sensitivity have put this method into focus, but the situation may
change considerably the coming few years as the CERN LHC experiments
collect data, and new gamma-ray detectors are being planned, such as
the CTA \cite{Consortium:2010bc}. In fact, it has recently been
pointed out \cite{Bergstrom:2010gh} that a dedicated ground-based
gamma-ray detector would have potential that goes far beyond that of
the other methods, depending on presently unknown parameters in the
particle physics models for DM.

Among the astrophysical systems which will be very interesting to detect, and
study, with gamma-ray detectors (\Fermi, H.E.S.S., MAGIC, VERITAS, and
eventually large detectors like CTA)\footnote{{High Energy Stereoscopic System,
    Major Atmospheric Gamma Imaging Cerenkov Telescope, Very Energetic Radiation
    Imaging Telescope Array System, Cerenkov Telescope Array.}} belong galaxy
clusters. The most promising directions in which to search for a gamma-ray
annihilation signal (from the annihilation process itself, and also the
accompanying bremsstrahlung and inverse Compton (IC) components coming from
charged particles produced in the annihilations) are basically three:

1. The galactic center (g.c.). This is where all numerical simulations of cold
dark matter (CDM) predict the highest density. However, the detailed DM density
in the very central part is difficult to predict, due a to a possibly very
complicated interplay between baryons, DM, and the central galactic black
hole. Also, it is a very crowded region with many gamma-ray sources like
pulsars, CR-illuminated molecular clouds, and other supernova remnants, which
have to be subtracted from the data to extract the DM induced signal. In fact,
there is a recent claim of an indication of a relatively light DM particle
contribution to the gamma-ray flux from the g.c. \cite{2010arXiv1010.2752H}, but
other hypotheses seem to work at least as well \cite{2010arXiv1012.5839B}.

2. The dwarf spheroidal galaxies orbiting the Milky Way (MW), like Segue-1, Ursa
Minor, Draco, Sagittarius, Sculptor, Carina or Willman-1
\cite{2009JCAP...01..016B,2010ApJ...720.1174A,2010JCAP...01..031S,2010JCAP...01..031S,2011arXiv1103.0477T,2011APh....34..608H}. The
problem here is that the nature of many of these small, dark matter-dominated
galaxies is not entirely clear, and the velocity dispersion estimates are based
on rather small numbers of stars.  Confusion with star clusters and tidal
disruption are other complications. Once a satellite dwarf galaxy is accreted by
the MW, the outer regions are severely affected by tidal stripping. The longer a
satellite has been part of our Galaxy, and the closer it comes to the center
during its pericentral passage, the more material is removed
\cite{2004MNRAS.355..819G}.  Thus the DM density profile is very uncertain for
most of them, especially for radii larger than those probed by the
stars. Nonetheless, by stacking the data together from many dwarf spheroidals
these uncertainties can be made less severe, and preliminary results from
\Fermi-LAT shows this method to give quite promising results \cite{garde}.

%For a satellite dwarf galaxy, however, once it is accreted by the MW,
%the outer regions are severely affected by tidal stripping. The longer
%a satellite has been part of our Galaxy, and the closer it comes to
%the center during its pericentral passage, the more material is
%removed \cite{2004MNRAS.355..819G}.

3. Galaxy clusters. This possibility has been less studied
theoretically, however currently there is an ongoing observational
campaign to detect gamma-ray emission from galaxy clusters
\cite{2003ApJ...588..155R,2006ApJ...644..148P,2008AIPC.1085..569P,2009A&A...495...27A,2009arXiv0907.0727T,2009arXiv0907.5000G,2009IJMPD..18.1627D,2009A&A...495...27A,2009A&A...502..437A,2009ApJ...704..240K,2009ApJ...706L.275A,2010ApJ...710..634A,2010JCAP...05..025A,2010ApJ...717L..71A}.
In fact, we noted in a previous Letter \cite{2009PhRvL.103r1302P} that
there are certain advantages that work in favour of this possible
target for gamma-ray detection of DM annihilation. Galaxy clusters
constitute the most massive objects in our Universe that are forming
today. This causes their DM subhalo mass function to be less affected
by tidal stripping compared to galaxy sized halos that formed long
ago. The annihilation luminosities of the DM halo component for
e.g. the Virgo cluster and the Draco dwarf scales in a way (see
\cite{2009PhRvL.103r1302P}) that the ratio of gamma-ray luminosities
from the smooth components is around 4, in favour of Virgo. In
addition, there may be a further enhancement due to substructure,
which to a large extent should be unaffected by tidal disruption, at
least in the outer regions. According to a recent estimate
\cite{2011MNRAS.410.2309G}, more massive haloes tend to have a larger
mass fraction in subhalos. For example, cluster size haloes typically
have 7.5 per cent of the mass within $r_{200}$ in substructures of
fractional mass larger than $10^{-5}$, which is 25 per cent higher
than for galactic haloes.\footnote{All halo masses and length scales
  in this paper are scaled to the currently favored value of Hubble's
  constant, $H_0 = 70\, \rmn{km~s}^{-1}\,\rmn{Mpc}^{-1}$. We define
  the virial mass $\mvir$ and virial radius $\rvir$ as the mass and
  radius of a sphere enclosing a mean density that is 200 times the
  critical density of the Universe $\rho_{\rmn{cr}}$.}

In this paper, we will investigate in detail the potential of several
of the most promising galaxy clusters to produce an annihilation
gamma-ray yield which could be observable with present and planned
gamma-ray detectors. Here, we sketch out the main arguments that need
to be considered for maximizing the expected signal-to-noise ratio of
a promising target cluster. First, we need to maximize the DM
annihilation flux of an unresolved cluster, $F=A_\rmn{dm} \int dV
\rho^2 / D^2 \sim \mvir^\alpha/D^2$, where $A_\rmn{dm}$ depends on the
particle physics model of DM, $\rho$ is the smooth DM density profile,
$D$ is the luminosity distance, and $\alpha=1$ if we assume
universality of the DM density profile. The dependence of $\rho$ on
the halo formation epoch breaks the universality and slightly modifies
the mass dependence, yielding $\alpha=0.83$
\cite{2009PhRvL.103r1302P}. Additionally, the presence of a hierarchy
of substructures down to small scales \cite{2008Natur.456...73S} and
the potential dependence of the particle physics cross section on the
relative DM particle velocities \cite{ArkaniHamed:2008qn} may
furthermore modify the scaling parameter $\alpha$ and shall be one of
the focus points of this work. Second, we need to minimize the
expected noise which is a sum of galactic and cluster-intrinsic
foreground. While the galactic foreground varies across the sky, it is
typically lower for increasing latitude (away from the galactic
plane). It is thought that the gamma-ray foreground from clusters is
dominated by CRs which are accelerated at cosmological formation
shocks and transported over time although there could be a substantial
contribution of CRs from AGNs and supernova-driven galactic winds. In
this work we will also study (and constrain) the CR-induced emission
from clusters to identify those objects that are expected to be
especially dim which would imply a low CR-induced background.  Since
the CR-induced emission is expected to scale with the thermal X-ray
emission of clusters \cite{2004A&A...413...17P}, we estimate gamma-ray
fluxes of a nearby X-ray flux complete sample of galaxy clusters from
the Rosat all-sky survey (extended HIghest X-ray FLUx Galaxy Cluster
Sample, HIFLUGCS,
\cite{2002ApJ...567..716R,2007A&A...466..805C}).\footnote{Note that we
  have added the Virgo cluster to the sample that we nevertheless
  refer to as extended HIFLUGCS catalogue in the following.}  This
complements and extends previous work related to DM in clusters
\cite{2006A&A...455...21C,2009PhRvD..80b3005J,2011arXiv1104.3530S,2011ApJ...726L...6C}
and to CRs in clusters
\cite{2010MNRAS.409..449P,2011MNRAS.410..127B,2008MNRAS.385.1211P,2009JCAP...08..002K,2010MNRAS.407.1565D}.

In Section 2 we discuss the theory of DM and gamma-rays. In
particular, we focus on the leptophilic (LP) and supersymmetric
benchmark (BM) DM models in this work, and outline the framework for
estimating the gamma-ray emission from various radiative processes. In
Section 3 we calculate the gamma-ray fluxes and spectral distributions
of four clusters that are identified as prime targets for DM
observations (Fornax and Virgo) as well as to have a high CR-induced
gamma-ray yield (Perseus and Coma). For the same clusters we derive
the gamma-ray surface brightness profiles in Section 4. To extend the
analysis to more clusters and increase the list of promising targets we
estimate the fluxes from DM and CRs of all clusters in the extended
HIFLUGCS sample in Section 5. In Section 6 we conclude and discuss our
results.

% In this paper, we will investigate in some more detail the potential
% of several of the most promising galaxy clusters to produce an
% annihilation gamma-ray yield which could be observable with present
% and planned gamma-ray detectors. For convenience, we summarize in
% Table~\ref{tab:flux_tab} our results for the gamma-ray flux from four
% promising clusters with a high gamma-ray yield. We will in the
% following derive the emission from two different DM models and a
% hadronic CR model to show that Virgo has the highest gamma-ray flux
% from annihilating DM. However, if the boost from Sommerfeld
% enhancement (SFE) is small then the gamma-ray emission induced by CRs
% is dominating most clusters with the exception of a few clusters; most
% notably Fornax that shows the largest ratio of DM annihilation-to-CR
% flux. In particular, we identify Fornax and Virgo as prime targets for
% DM observations, Perseus for CR induced emission and we additionally
% decide in favor of the well studied cluster Coma for comparison. While
% studying these four clusters in great detail, we complement the
% analysis by calculating the expected gamma-ray fluxes of a nearby
% X-ray flux complete sample of galaxy clusters from the Rosat all-sky
% survey (extended HIghest X-ray FLUx Galaxy Cluster Sample, HIFLUGCS,
% \cite{2002ApJ...567..716R,2007A&A...466..805C}).\footnote{Note that
%   we have added the Virgo cluster to the sample that we nevertheless
%   refer to as extended HIFLUGCS catalogue in the following.} For
% previous work related to DM in clusters, see, e.g.,
% \cite{2006A&A...455...21C,2009PhRvD..80b3005J,2011arXiv1104.3530S,2011ApJ...726L...6C,2010JCAP...05..025A}
% and for recent work on CRs in clusters see
% \cite{2010MNRAS.409..449P,2011MNRAS.410..127B,2008MNRAS.385.1211P,2009JCAP...08..002K,2010MNRAS.407.1565D}.

@@

There are three methods for detecting DM candidates which are
presently employed: First, at particle physics accelerators such as
CERN's LHC, one is now entering into a new energy regime which may
allow the production of the heavy, electrically neutral and long-lived
particles which may constitute DM. From these experiments one may get
the first glimpse of the mass scale beyond the standard model where DM
may reside. However, to really show that any of the hypothetical, new
particles created at the LHC is actually the DM, one has to rely on
the two other methods available for the DM search, namely direct and
indirect detection. Direct detection methods, which are presently
evolving rapidly use the feeble interaction between DM particles of
the galactic halo and nuclei such as Germanium, Sodium, Iodine, Argon
or Xenon to infer the scattering cross section and a rough estimate of
the mass of the DM particle (for the currently most sensitive search,
see \cite{Aprile:2010um,Aprile:2011hi}). A characteristic of this
method is that it basically only depends on the local DM density
(which is rather well determined to be around $0.4$ GeV/cm$^3$) and
the basic cross section for DM - nucleus scattering. A disadvantage is
that it does not benefit from the two enhancement mechanisms mentioned
above, boost from substructure and/or the Sommerfeld effect. Also, it
cannot be excluded - although it seems at present improbable - that
the local halo structure is such that the solar system happens to be
in an underdense region.

In indirect detection, on the other hand, in particular in the photon
channel, one searches for products of DM annihilation in the galactic
halo and beyond. Particularly interesting targets are the galactic
center (for a recent possible indication of a signal, see
\cite{2010arXiv1010.2752H}, however, see also
\cite{2010arXiv1012.5839B}), the dwarf galaxies surrounding the MW
\cite{Strigari:2006rd,Essig:2009jx,2010JCAP...01..031S}, galaxy
clusters (the focus of this work)
\cite{Ghigna:1998vn,Lewis:2002mfa,Boyarsky:2006kc,2006A&A...455...21C,2009PhRvL.103r1302P},
and even the cosmological large scale structure
\cite{Bergstrom:2001jj,Ullio:2002pj,Taylor:2002zd,Elsaesser:2004ap,2011MNRAS.tmp..503C,Abazajian:2010sq,Abdo:2010dk,Zavala:2011tt}.

>>>>

%NEW
\newcommand{\rmn}{\mathrm}
\newcommand{\bx}{{\bf x}}
\newcommand{\fg}{{F_\gamma}}
\newcommand{\sg}{{S_\gamma}}
\newcommand{\psf}{\theta_\rmn{res}}
\newcommand{\ph}{\rmn{ph}}
\newcommand{\eph}{E_\ph}
\newcommand{\nph}{N_\ph}
\newcommand{\vir}{\rmn{vir}}
\newcommand{\gal}{\rmn{gal}}
\newcommand{\sd}{\rmn{SD}}
\newcommand{\clu}{\rmn{cl}}
\newcommand{\sfe}{\rmn{sfe}}
\newcommand{\sub}{\rmn{sub}}
\newcommand{\msun}{M_\odot}
\newcommand{\ee}{E_\rmn{e}}
\newcommand{\stars}{\rmn{stars}}
\newcommand{\dust}{\rmn{dust}}
\newcommand{\lx}{L_\rmn{X}}
\newcommand{\s}{\rmn{s}}
\newcommand{\Kp}{\rmn{K}^\prime}
\newcommand{\Ip}{\rmn{I}^\prime}
\newcommand{\Js}{\rmn{J}^*}
\newcommand{\Jp}{\rmn{J}^\prime}
\newcommand{\B}{\rmn{B}}
\newcommand{\bsub}{\B_\rmn{sub}}
\newcommand{\qCR}{q_{\rmn{CR}-\ensuremath{\pi^0}}}
\newcommand{\ds}{{\sc DarkSUSY}}
\newcommand{\mlim}{M_\rmn{lim}}
\newcommand{\sm}{\rmn{sm}}
\newcommand{\mpc}{\rmn{Mpc}}
\newcommand{\kpc}{\rmn{kpc}}
\newcommand{\mev}{\rmn{MeV}}
\newcommand{\cm}{\rmn{cm}}
\newcommand{\egt}{\tilde{E}_\gamma}
\newcommand{\eet}{\tilde{E}_\rmn{e}}
\newcommand{\bra}{\langle}
\newcommand{\ket}{\rangle}
\newcommand{\vst}{\vspace{-0.14mm}}
\newcommand{\vstt}{\vspace{-0.0mm}}
\newcommand{\el}{\rmn{el}}
\newcommand{\mytimes}{\hspace{-0.5mm}\times\hspace{-0.5mm}}
\newcommand{\hst}{\hspace{-1.0mm}}

@@

\begin{abstract}
  \C{We study the possibility for detecting gamma-ray emission from
  galaxy clusters. We consider 1) leptophilic models of dark matter
  (DM) annihilation that include a Sommerfeld enhancement (SFE), 2)
  different representative benchmark models of supersymmetric DM, and
  3) cosmic ray (CR) induced pion decay. To model DM annihilation, we
  consider continuum emission as a result of the hadronization of
  annihilating neutralinos, internal bremsstrahlung, and inverse
  Compton emission from the cosmic microwave background as well as
  from a realistic spatial and spectral distribution of dust and
  stellar light. Among all clusters/groups of a flux-limited X-ray
  sample, we predict Virgo, Fornax and M49 to be the brightest DM
  sources and find a particularly low CR-induced background for
  Fornax. For a minimum substructure mass given by the DM
  free-streaming scale, cluster halos maximize the substructure boost
  for which we find a factor of $\gtrsim 1000$. Since regions around
  the virial radius dominate the annihilation flux of substructures,
  the resulting surface brightness profiles are almost flat. This
  makes it very challenging to detect this flux with imaging
  atmospheric Cherenkov telescopes since their sensitivity drops
  approximately linearly with radius and they typically have $5-10$
  linear resolution elements across a cluster. Assuming cold dark
  matter with a substructure mass distribution down to an Earth mass
  and using extended \Fermi upper limits, we rule out the leptophilic
  models in their present form in 28 clusters, and limit the boost
  from SFE in M49 and Fornax to be $\lesssim 5$. This corresponds to a
  limit on SFE in the Milky Way of $\lesssim 3$ which is too small to
  account for the increasing positron fraction with energy as seen by
  PAMELA. Alternatively, if SFE is realized in Nature, this would
  imply a limiting substructure mass of $\mlim > 10^4\,\msun$---a
  problem for structure formation in most particle physics
  models. Using individual cluster observations, it will be
  challenging for \Fermi to constrain our selection of DM benchmark
  models without SFE. The \Fermi upper limits are, however, closing in
  on our predictions for the CR flux using an analytic model based on
  cosmological hydrodynamical cluster simulations. We limit the
  CR-to-thermal pressure in nearby bright galaxy clusters of the
  \Fermi sample to $\lesssim 10\%$ and in Norma and Coma to $\lesssim
  3\%$. Thus we will soon start to constrain the underlying CR physics
  such as shock acceleration efficiencies or CR transport properties.}
\end{abstract}

@@

\C{Dark matter (DM) has been searched for in direct detection
  experiments \cite{Pato:2010zk}, at accelerators
  \cite{Ellis:2001hv,Baer:2006ff,Khachatryan:2011tk} and also in
  indirect detection experiments looking for signals in the cosmic-ray
  (CR) spectra of antiprotons, positrons, neutrinos and all of the
  electromagnetic spectrum from radio waves to gamma-rays
  \cite{Bergstrom:2009ib}. So far, the improvements in direct
  detection sensitivity have put this method into focus, but the
  situation may change considerably the coming few years as the CERN
  LHC experiments collect data, and new gamma-ray detectors are being
  planned, such as the CTA \cite{Consortium:2010bc}. In fact, it has
  recently been pointed out \cite{Bergstrom:2010gh} that a dedicated
  ground-based gamma-ray detector would have potential that goes far
  beyond that of the other methods, depending on presently unknown
  parameters in the particle physics models for DM.}

\C{Among the astrophysical systems which will be very interesting to
  detect, and study, with gamma-ray detectors (\Fermi, H.E.S.S.,
  MAGIC, VERITAS, and eventually large detectors like
  CTA)\footnote{{High Energy Stereoscopic System, Major Atmospheric
      Gamma Imaging Cerenkov Telescope, Very Energetic Radiation
      Imaging Telescope Array System, Cerenkov Telescope Array.}}
  belong galaxy clusters. The most promising directions in which to
  search for a gamma-ray annihilation signal (from the annihilation
  process itself, and also the accompanying bremsstrahlung and inverse
  Compton (IC) components coming from charged particles produced in
  the annihilations) are basically three:}

\C{1. The galactic center (g.c.). This is where all numerical
  simulations of cold dark matter (CDM) predict the highest
  density. However, the detailed DM density in the very central part
  is difficult to predict, due a to a possibly very complicated
  interplay between baryons, DM, and the central galactic black
  hole. Also, it is a very crowded region with many gamma-ray sources
  like pulsars, CR-illuminated molecular clouds, and other supernova
  remnants, which have to be subtracted from the data to extract the
  DM induced signal. In fact, there is a recent claim of an indication
  of a relatively light DM particle contribution to the gamma-ray flux
  from the g.c. \cite{2011PhLB..697..412H}, but other hypotheses seem
  to work at least as well \cite{2010arXiv1012.5839B}.}

\C{2. The dwarf spheroidal galaxies orbiting the Milky Way (MW), like
  Segue-1, Ursa Minor, Draco, Sagittarius, Sculptor, Carina or
  Willman-1
  \cite{2009JCAP...01..016B,2010ApJ...720.1174A,2010JCAP...01..031S,2010JCAP...01..031S,2011arXiv1103.0477T,2011APh....34..608H}. The
  problem here is that the nature of many of these small, dark
  matter-dominated galaxies is not entirely clear, and the velocity
  dispersion estimates are based on rather small numbers of stars.
  Confusion with star clusters and tidal disruption are other
  complications. Once a satellite dwarf galaxy is accreted by the MW,
  the outer regions are severely affected by tidal stripping. The
  longer a satellite has been part of our Galaxy, and the closer it
  comes to the center during its pericentral passage, the more
  material is removed \cite{2004MNRAS.355..819G}.  Thus the DM density
  profile is very uncertain for most of them, especially for radii
  larger than those probed by the stars. Nonetheless, by stacking the
  data together from many dwarf spheroidals these uncertainties can be
  made less severe, and preliminary results from \Fermi-LAT shows this
  method to give quite promising results \cite{garde}.}

\C{3. Galaxy clusters. This possibility has been less studied
  theoretically, however currently there is an ongoing observational
  campaign to detect gamma-ray emission from galaxy clusters
  \cite{2003ApJ...588..155R,2006ApJ...644..148P,2008AIPC.1085..569P,2009A&A...495...27A,2009arXiv0907.5000G,2009IJMPD..18.1627D,2009A&A...495...27A,2009A&A...502..437A,2009ApJ...704..240K,2009ApJ...706L.275A,2010ApJ...710..634A,2010JCAP...05..025A,2010ApJ...717L..71A}.
  In fact, we noted in a previous Letter \cite{2009PhRvL.103r1302P}
  that there are certain advantages that work in favour of this
  possible target for gamma-ray detection of DM annihilation. Galaxy
  clusters constitute the most massive objects in our Universe that
  are forming today. This causes their DM subhalo mass function to be
  less affected by tidal stripping compared to galaxy sized halos that
  formed long ago. The annihilation luminosities of the DM halo
  component for e.g. the Virgo cluster and the Draco dwarf scales in a
  way (see \cite{2009PhRvL.103r1302P}) that the ratio of gamma-ray
  luminosities from the smooth components is around 4, in favour of
  Virgo. In addition, there may be a further enhancement due to
  substructure, which to a large extent should be unaffected by tidal
  disruption, at least in the outer regions. According to a recent
  estimate \cite{2011MNRAS.410.2309G}, more massive haloes tend to
  have a larger mass fraction in subhalos. For example, cluster size
  haloes typically have 7.5 per cent of the mass within $r_{200}$ in
  substructures of fractional mass larger than $10^{-5}$, which is 25
  per cent higher than for galactic haloes.\footnote{All halo masses
    and length scales in this paper are scaled to the currently
    favored value of Hubble's constant, $H_0 = 70\,
    \rmn{km~s}^{-1}\,\rmn{Mpc}^{-1}$. We define the virial mass
    $\mvir$ and virial radius $\rvir$ as the mass and radius of a
    sphere enclosing a mean density that is 200 times the critical
    density of the Universe $\rho_{\rmn{cr}}$.}}

\C{In this paper, we will investigate in detail the potential of
  several of the most promising galaxy clusters to produce an
  annihilation gamma-ray yield which could be observable with present
  and planned gamma-ray detectors. Here, we sketch out the main
  arguments that need to be considered for maximizing the expected
  signal-to-noise ratio of a promising target cluster. First, we need
  to maximize the DM annihilation flux of an unresolved cluster,
  $F=A_\rmn{dm} \int dV \rho^2 / D^2 \sim \mvir^\alpha/D^2$, where
  $A_\rmn{dm}$ depends on the particle physics model of DM, $\rho$ is
  the smooth DM density profile, $D$ is the luminosity distance, and
  $\alpha=1$ if we assume universality of the DM density profile. The
  dependence of $\rho$ on the halo formation epoch breaks the
  universality and slightly modifies the mass dependence, yielding
  $\alpha=0.83$ \cite{2009PhRvL.103r1302P}. Additionally, the presence
  of a hierarchy of substructures down to small scales
  \cite{2008Natur.456...73S} and the potential dependence of the
  particle physics cross section on the relative DM particle
  velocities \cite{ArkaniHamed:2008qn} may furthermore modify the
  scaling parameter $\alpha$ and shall be one of the focus points of
  this work. Second, we need to minimize the expected noise which is a
  sum of galactic and cluster-intrinsic foreground. While the galactic
  foreground varies across the sky, it is typically lower for
  increasing latitude (away from the galactic plane). It is thought
  that the gamma-ray foreground from clusters is dominated by CRs
  which are accelerated at cosmological formation shocks and
  transported over time although there could be a substantial
  contribution of CRs from AGNs and supernova-driven galactic
  winds. In this work we will also study (and constrain) the
  CR-induced emission from clusters to identify those objects that are
  expected to be especially dim which would imply a low CR-induced
  background.  Since the CR-induced emission is expected to scale with
  the thermal X-ray emission of clusters \cite{2004A&A...413...17P},
  we estimate gamma-ray fluxes of a nearby X-ray flux complete sample
  of galaxy clusters from the Rosat all-sky survey (extended HIghest
  X-ray FLUx Galaxy Cluster Sample, HIFLUGCS,
  \cite{2002ApJ...567..716R,2007A&A...466..805C}).\footnote{Note that
    we have added the Virgo cluster to the sample that we nevertheless
    refer to as extended HIFLUGCS catalogue in the following.}  This
  complements and extends previous work related to DM in clusters
  \cite{2006A&A...455...21C,2009PhRvD..80b3005J,2011arXiv1104.3530S,2011ApJ...726L...6C,2011arXiv1110.1529H}
  and to CRs in clusters
  \cite{2010MNRAS.409..449P,2011MNRAS.410..127B,2008MNRAS.385.1211P,2009JCAP...08..002K,2010MNRAS.407.1565D}.}

\C{In Section 2 we discuss the theory of DM and gamma-rays. In
particular, we focus on the leptophilic (LP) and supersymmetric
benchmark (BM) DM models in this work, and outline the framework for
estimating the gamma-ray emission from various radiative processes. In
Section 3 we calculate the gamma-ray fluxes and spectral distributions
of four clusters that are identified as prime targets for DM
observations (Fornax and Virgo) as well as to have a high CR-induced
gamma-ray yield (Perseus and Coma). For the same clusters we derive
the gamma-ray surface brightness profiles in Section 4. To extend the
analysis to more clusters and increase the list of promising targets we
estimate the fluxes from DM and CRs of all clusters in the extended
HIFLUGCS sample in Section 5. In Section 6 we conclude and discuss our
results.}

@@

There are three methods for detecting DM candidates which are
presently employed: First, at particle physics accelerators such as
CERN's LHC, one is now entering into a new energy regime which may
allow the production of the heavy, electrically neutral and long-lived
particles which may constitute DM. From these experiments one may get
the first glimpse of the mass scale beyond the standard model where DM
may reside. However, to really show that any of the hypothetical, new
particles created at the LHC is actually the DM, one has to rely on
the two other methods available for the DM search, namely direct and
indirect detection. \C{Second,} direct detection methods, which are
presently evolving rapidly, use the feeble interaction between DM
particles of the galactic halo and nuclei such as Germanium, Sodium,
Iodine, Argon or Xenon to infer the scattering cross section and a
rough estimate of the mass of the DM particle (for the currently most
sensitive search, see \cite{Aprile:2010um,Aprile:2011hi}). A
characteristic of this method is that it basically only depends on the
local DM density (which is rather well determined to be around $0.4$
GeV/cm$^3$) and the basic cross section for DM - nucleus scattering. A
disadvantage is that it does not benefit from the two enhancement
mechanisms mentioned above, boost from substructure and/or the
Sommerfeld effect. Also, it cannot be excluded - although it seems at
present improbable - that the local halo structure is such that the
solar system happens to be in an underdense region.

\C{Third}, in indirect detection, \del{on the other hand,} in
particular in the photon channel, one searches for products of DM
annihilation in the galactic halo and beyond. Particularly interesting
targets are the galactic center (for a recent possible indication of a
signal, see \cite{2011PhLB..697..412H}, however, see also
\cite{2010arXiv1012.5839B}), the dwarf galaxies surrounding the MW
\cite{Strigari:2006rd,Essig:2009jx,2010JCAP...01..031S}, galaxy
clusters (the focus of this work)
\cite{Ghigna:1998vn,Lewis:2002mfa,Boyarsky:2006kc,2006A&A...455...21C,2009PhRvL.103r1302P},
and even the cosmological large scale structure
\cite{Bergstrom:2001jj,Ullio:2002pj,Taylor:2002zd,Elsaesser:2004ap,2011MNRAS.tmp..503C,Abazajian:2010sq,Abdo:2010dk,Zavala:2011tt}.

\subsubsection{Leptophilic models}
\label{sect:LP}
\C{There is also a possibility to search for DM annihilation in the Milky
Way indirectly through annihilation to antimatter channels, which
however lacks the important directional signature of
gamma-rays. Recently, it has however been much in focus due to the
surprising findings of PAMELA \cite{Adriani:2008zr} and \Fermi
\cite{Abdo:2009zk}. Viable particle physics models are rather
constrained by other observations, however. For example, the
non-observation of an enhanced antiproton flux by PAMELA
\cite{Adriani:2010rc} means that quark-antiquark final states have to
be suppressed. This has led to the postulate of a whole class of
{\em leptophilic} models with mainly or exclusively annihilation to
leptons and antileptons. In addition, one has to have a sizeable
fraction to final states containing muons or tau leptons as the shape
of the spectrum disfavors direct annihilation to electrons and
positrons only. A large enhancement, of the order of at least several
hundred, of the annihilation rate is also needed, something that may
be given by the {\em Sommerfeld effect} (for a pedagogical review of
viable scenarios, see \cite{ArkaniHamed:2008qn}). As still another
difficulty for these models, the lack of an IC signature from the
Galactic center means that the DM density distribution must be cored
rather than cuspy
\cite{Bertone:2008xr,Cirelli:2008pk,Bergstrom:2008ag}. (For a review
of the DM modeling of these effects, see \cite{Bergstrom:2009ib}. For
a recent treatment, showing still viable models, see
\cite{Finkbeiner:2010sm}.)}

\C{A more standard explanation of the PAMELA/\Fermi excess would be
pulsars or other supernova remnants (e.g.,
\cite{1995PhRvD..52.3265A,Hooper:2008kg,Ahlers:2009ae}). In that case,
DM would more likely be explained by the conventional scenario of a
WIMP, for example the thoroughly studied lightest supersymmetric
neutralino (for reviews, see
\cite{Jungman:1995df,Bergstrom:2000pn,Bertone:2004pz}). We note that
AMS-02 experiment on the International Space Station \cite{ams02} may give
interesting clues to the origin of the antimatter excess.}
